Supporting Information:\\Black-swan events in animal populations

### 2016-05-27 text/ms.tex
Add caveats and modeling suggestion paragraphs

Given the multitude of reasons why a population modeler might expect heavy tail process noise, what should a population modeler do?

Essentially any reason for why a population modeler has incorrectly specified
the exact process generating the abundance times there is under question could
generate have detailed process noise.

We think modelers should routinely consider substituting a heavy tailed
distribution, such as the student t-distribution, to represent the process
noise in population dynamic modeling. Especially when making forecasts to make
predictions about risk of specified bad events.

But what heavy tailed distribution to pick?

This is an open research question. However, we think the student t-distribution
represents a simple solution in that it can be easily substituted for the
normal distribution and if the degrees of freedom are fixed at a moderately
heavy level it doesn't require the estimation of any extra parameters over the
normal distribution.

And if we use a student t-distribution than what degrees of freedom parameter should be you assume?

The exact shape of the tails has been shown to be important in the field of
dispersal biology -- has a big effect on dynamic \citep{kot1996}
\citep{clark1999}
\citep{kot1996}

The degrees of
freedom could be estimated from meta-analysis but even here using one of the
best large data sets of animal abundance timeseries the timeseries are
relatively short and the taxonomy is a biased sample of species of interest to humans  economic
interest ...

for example the majority of the blocks one offense that we observed were
population crashes This could be due to basic biology --- animal birthrates are
constrained whereas every animal can die --- on the other hand it could be due
to the biased sample of species in the data set which represent species of
concern to humans and we don't have much data for highly reproductive species
that are not threatened or of economic interest

If there is any reason for a model are to expect that external forces could act
or that there is a mixture of distributions through spatial heterogeneity we
think it is prudent to err on the side of caution and allow for heavy tailed
process noise in forecasts

The barring a better estimate we might suggest using a degrees of freedom parameter around 5

this allows for a catastrophic event of an x fold decline of abundance that
happens every Y years compared to a normal distribution allowing only for this
event to happen every Z years.

But what was once considered a one in X million event then becomes a one in Y
million event

A t-distribution compared to a normal distribution on the process error he
essentially lets us be right most of the time while allowing are models to be
very wrong occasionally this better matches our current intuition that
ecological surprises can and do occur

Our hypothesis is that our first priority is to allow for tails that are
heavier than the normal distribution and approximately in line with those
discovered in this paper and our second priority is to determine exactly how
heavy a default distribution we should assume.

Answers might come from simulation analyses exploring the implications of
different tail shapes and also from fitting population models to known time
series of mass mortality events \citep[e.g.][]{fey2015}

>>>>

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focus on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Additionally, some apparent black-swan events could be
recording mistakes, or conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors. Finally, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.

We think modelers should routinely consider representing population dynamic
process noise with a (possibly downward-skewed)
heavy-tailed distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to timeseries
with suspected mass mortality events
\citep[e.g.~the t distribution compared to a mixture
of two normal distributions;][]{ward2007}.
In the meantime, we think the t distribution
represents a simple solution.
It can be easily substituted for the normal distribution
and if the degrees of freedom parameter, $\nu$,
is fixed
it requires estimating no more parameters
then the normal distribution.
Barring a better estimate, we suggest $\nu = 4$ or $5$ and possibly a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a).
Above $\nu  =  10$ the t distribution closely resembles
the normal distribution,
and $\nu  = 4$ or $5$ allows for an event
three standard deviations below the mean to happen
once every 50--65 years instead of once every $\sim$ 740 years.
Adopting heavy-tailed process noise follows the intuition
that our models are usually reasonable but occasionally very wrong.
